---
title: "Single_Cell_Cortex"
author: "Nischal Bhandari"
date: "7/11/2023"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
```{r}
library(Seurat)
```
Load the data
```{r}
count_matrix <- Matrix::readMM("filtered_count_matrix.mtx")
gene_names <- read.table("filtered_gene_row_names.txt", header = FALSE)
meta_data <- read.table("filtered_column_metadata.txt", header = TRUE)
rownames(meta_data) <- meta_data$TAG
```

Name the rows and colmnames
```{r}
dimnames(count_matrix) <- list(gene_names$V1, meta_data$TAG)
```


Create the seurat object for cortex single cell data.
```{r}
cortex_single <- CreateSeuratObject(counts = count_matrix, min.cells = 3, min.features = 300)
cortex_single[["percent.mt"]] <- PercentageFeatureSet(cortex_single, pattern = "^MT-")
VlnPlot(cortex_single, features = c("nFeature_RNA", "nCount_RNA", "percent.mt"), ncol = 3, pt.size = 0.1) + NoLegend()
```

Add the metadata

```{r}
cortex_single <- AddMetaData(cortex_single, metadata = meta_data)
```

```{r}
# variable features
cortex_single <- NormalizeData(cortex_single)
cortex_single <- FindVariableFeatures(cortex_single, selection.method = "vst", nfeatures = 2000)

# scale the data
all.genes <- rownames(cortex_single)
cortex_single <- ScaleData(cortex_single) #,features = all.genes)

# RUN pcas
cortex_single <- RunPCA(cortex_single, features = VariableFeatures(object = cortex_single))
VizDimLoadings(cortex_single, dims = 1:2, reduction = "pca")

#plot pcas
DimPlot(cortex_single, reduction = "pca", group.by = 'broad.cell.type')

```
Find neighbors
```{r}
cortex_single <- FindNeighbors(cortex_single, dims = 1:50)

# Run UMAP
cortex_single <- RunUMAP(cortex_single, dims = 1:50)

# Plot UMAPs
p1 <- DimPlot(cortex_single, reduction = "umap", group.by = 'Subcluster')

# save the RDS
saveRDS(cortex_single, "cortex_single_seurat.rds")
cortex_single <- readRDS("cortex_single_seurat.rds")
```

```{r}
# select_type <- c("Ast", "End", "Ex", "In", "Mic", "Oli")
saveRDS(cortex_single, "cortex_single_seurat.rds")

```

Run Deconvolution Using scCortex Sample and MuSiC

```{r}
library(MuSiC)

library(SummarizedExperiment)
library(tidyr)
library(dplyr)
library(ggplot2)
```

```{r}
# relable the short names
# merge cells into neurons
cortex_single@meta.data <- cortex_single@meta.data %>%
  mutate(broad.cell.type = case_when(
     broad.cell.type == "Oli" ~ "Oligodendrocytes",
    broad.cell.type %in% c("In", "Ex") ~ "Neurons",
    broad.cell.type == "Mic" ~ "Microglia",
    broad.cell.type == "Ast" ~ "Astrocytes",
    broad.cell.type == "End" ~ "Endothelial",
    broad.cell.type == "Per" ~ "Pericytes",
    TRUE ~ broad.cell.type 
  ))
```

```{r}
count_cell_type <- cortex_single@meta.data %>%
  group_by(broad.cell.type) %>%
  summarize(count = n())
count_cell_type[order(count_cell_type$count, decreasing = TRUE), ]
```

```{r}
# changing the single cell SO to SCE
singlecortex.sce <- as.SingleCellExperiment(cortex_single)
```
Reading bulk data

```{r}

all_bulk_samples <- load("/bulk_matrix.RData")
all_bulk_samples <- genes_counts
all_bulk_samples$gene <- rownames(all_bulk_samples)
all_bulk_samples$gene <- sub("\\..*$", "", all_bulk_samples$gene)

all_bulk_samples <- all_bulk_samples %>% filter(!duplicated(all_bulk_samples$gene))



genes_names_df <- read.csv("genemes_names_df.csv")
genes_names_df$X <- NULL
genes_names_df <- na.omit(genes_names_df)
all_bulk_samples <- all_bulk_samples[match(genes_names_df$gene_ids, all_bulk_samples$gene), ]
all_bulk_samples <- na.omit(all_bulk_samples)
all_bulk_samples <- subset(all_bulk_samples, gene %in% genes_names_df$gene_ids)
genes_names_df <- genes_names_df[match(all_bulk_samples$gene, genes_names_df$gene_ids), ]

all_bulk_samples <- na.omit(all_bulk_samples)
all_bulk_samples$gene <- NULL


# getting the rownames

rownames(all_bulk_samples) <- genes_names_df$genes
all_bulk_samples <- data.matrix(all_bulk_samples)
any(is.na(all_bulk_samples))

#making expressionset

# for cortex data
final_spinalled_ematrix <- all_bulk_samples
write.csv(final_spinalled_ematrix, "bulk_cortex.csv")

colnames(all_bulk_samples)[1] <- "gene"
rownames(all_bulk_samples) <- all_bulk_samples$gene
all_bulk_samples$gene <- NULL
colnames(all_bulk_samples) <- gsub("[.]", "-", colnames(all_bulk_samples))

```

```{r}
meta_data <- read.csv("metada.csv")

als_mnd <- meta_data[meta_data$Subject.Group == "ALS Spectrum MND", ]
als_mnd_sample_names <- als_mnd$X

control <-  meta_data[meta_data$Subject.Group == "Non-Neurological Control", ]
control_sample_names <- control$X
```

Deoconvulate with MuSiC
```{r}
# MuSiC
bulk_control_healthy <- all_bulk_samples
bulk_control_healthy <- bulk_control_healthy[, colnames(bulk_control_healthy) %in% c(control_sample_names, als_mnd_sample_names)]
bulk_control_healthy <- bulk_control_healthy[, colnames(bulk_control_healthy) %in% cortex_samples]

bulk_control_healthy <- na.omit(bulk_control_healthy)
bulk_control_healthy <- as.matrix(bulk_control_healthy)
prop_music_cortex =music_prop(bulk.mtx = bulk_control_healthy
                      , sc.sce = singlecortex.sce,
                      clusters = 'broad.cell.type', samples = 'TAG', verbose = F)


saveRDS(prop_music_cortex, "CortexReferenceCortexEstimation_all.rds")

```
Plot the cell type estimations with jitter plot
```{r}
jitter.fig = Jitter_Est(list(data.matrix(prop_music$Est.prop.weighted),
                             data.matrix(prop_music$Est.prop.allgene)),
                        method.name = c('MuSiC', 'NNLS'), title = 'Jitter plot of Est Proportions')


```

```{r}
music_predicted <- prop_music$Est.prop.weighted
nnls_predicted <- prop_music$Est.prop.allgene


 
prop_all_m = cbind('proportion'=c(music_predicted),'celltype'=rep(colnames(music_predicted), each=nrow(music_predicted)), 'sampleID'=rep(rownames(music_predicted),times=ncol(music_predicted)),'Method'='MuSiC')
prop_all_n = cbind('proportion'=c(nnls_predicted),'celltype'=rep(colnames(nnls_predicted), each=nrow(nnls_predicted)), 'sampleID'=rep(rownames(nnls_predicted),times=ncol(nnls_predicted)),'Method'='NNLS')

prop_all_m=as.data.frame(prop_all_m)
prop_all_n=as.data.frame(prop_all_n)
prop_all_m$proportion=as.numeric(as.character(prop_all_m$proportion))
prop_all_n$proportion=as.numeric(as.character(prop_all_n$proportion))
ALL=rbind(prop_all_m, prop_all_n)
ALL$group = ifelse(ALL$sampleID %in% control_sample_names, 'Healthy', 'ALS')
```

Calculating P-Values

```{r}
# Subsetting to cortex samples alone

meta_data_cortex <- subset(meta_data, grepl("Cortex", Sample.Source, ignore.case = TRUE))
cortex_samples <- meta_data_cortex$X

ALL <- ALL[ALL$sampleID %in% cortex_samples, ]
ALL$p_nom <- pairwise.wilcox.test(x = ALL$proportion, g = ALL$group)$p.value[1]
ALL$p_resid <- pairwise.wilcox.test(x = ALL$proportion, g = ALL$group)$p.value[1]
```

Calculating per cell type

```{r}
ALL <- ALL[!ALL$celltype == 'Opc', ]
ALL <- ALL %>%
  mutate(celltype = case_when(
    celltype == "Oli" ~ "Oligodendrocytes",
    celltype %in% c("In", "Ex") ~ "Neurons",
    celltype == "Mic" ~ "Microglia",
    celltype == "Ast" ~ "Astrocytes",
    celltype == "End" ~ "Endothelial",
    celltype == "Per" ~ "Pericytes",
    TRUE ~ celltype 
  ))
```

```{r}
# linear model to account fo other factors
meta_data_cortex <- meta_data_cortex[match(ALL$sampleID, meta_data_cortex$X), ]

All_with_meta <- cbind(ALL, meta_data_cortex)
All_with_meta$X <- NULL

All_with_meta <- All_with_meta %>%
  mutate(Sample.Source = ifelse(grepl("^Cortex_Frontal", Sample.Source), "Cortex_Frontal", Sample.Source))
All_with_meta <- All_with_meta %>%
  mutate(Sample.Source = ifelse(grepl("^Cortex_Motor", Sample.Source), "Cortex_Motor", Sample.Source))
All_with_meta_NA <- na.omit(All_with_meta)
mod <- lm( proportion ~ factor(Site.Specimen.Collected) +  factor(Prep) + factor(Platform), All_with_meta)

All_with_meta$resid <-  residuals(mod)

All_with_meta$p_resid <- pairwise.wilcox.test(x = All_with_meta$resid, g = All_with_meta$group)$p.value[1]
```

```{r}
# Create an empty list to store the p-values for each cell type
p_nom <- list()
p_resid <- list()
# Get unique cell types from the 'celltype' column
unique_celltypes <- unique(All_with_meta$celltype)

# Iterate over each unique cell type
for (cell in unique_celltypes) {
  # Subset the data for the current cell type
  subset_data <- subset(All_with_meta, celltype == cell)
  
  # Perform pairwise Wilcoxon test for the current cell type
  pairwise_result <- pairwise.wilcox.test(x = subset_data$proportion, g = subset_data$group)
  residual_result <- pairwise.wilcox.test(x = subset_data$resid, g = subset_data$group)
  # Extract the p-values and store them in the list
  p_nom[[cell]] <- pairwise_result$p.value[1]
  p_resid[[cell]] <- residual_result$p.value[1]
}

All_with_meta$p_nom <- NULL
All_with_meta$p_resid <- NULL

# store the p-values
p_values_df <- data.frame(celltype = names(p_nom), p_nom = unlist(p_nom), p_resid = unlist(p_resid), stringsAsFactors = FALSE)

# Merge the p_values_df with the ALL data frame
All_with_meta <- merge(All_with_meta, p_values_df, by = "celltype", all.x = TRUE)

```

plotting the box_plot and p_Values

```{r}

library(ggplot2)

# the boxplot
plot <- ggplot(All_with_meta, aes(x = celltype, y = proportion, fill = group)) +
  geom_boxplot(outlier.colour = NA) +
  labs(x = "Cell Type", y = "Proportion", fill = "Group")

# Add p-values as text above each cell type
plot <- plot +
  geom_text(data = All_with_meta[!duplicated(ALL$celltype), ],
            aes(label = paste0("p = ", p_value)),
            position = position_dodge(width = 0.35), vjust = -20)

# Adjust the plot appearance
plot <- plot +
  theme_bw() +
  theme(legend.position = "right", 
        axis.text.x = element_text(angle = 45, hjust = 1),
        plot.title = element_text(face = "bold"))

# Display the plot
plot

```
```{r}

write.csv(All_with_meta, "final_cortex_estimations_with_p_values.csv")
```

Plotting in the style of Humphrey

```{r}
  
p_df <- All_with_meta %>% 
    group_by(celltype, Sample.Source) %>% 
    summarise(p_nom = min(p_nom), p_resid = min(p_resid)) %>%
    mutate(padj = p.adjust(p_resid, method = "bonferroni")) %>%
    mutate(padj_label = case_when(
    padj < 0.0001 ~ "***",
    padj < 0.001 ~ "**",
    padj < 0.05 ~ "*",
    TRUE ~ "."
    ))
P_df_subset <- p_df[, c("celltype", "padj", "padj_label", "Sample.Source")]
All_with_meta <- merge(All_with_meta, P_df_subset, by = c("celltype", "Sample.Source"), all.x = TRUE)
All_with_meta <- All_with_meta[All_with_meta$Method == "MuSiC", ]
```

Saving all faceted image

```{r}
plot_all <- All_with_meta[All_with_meta$Sample.Source %in% c("Cortex_Frontal", "Cortex_Motor"), ] %>%
  ggplot(aes(x = group, proportion)) +
      geom_point(aes(colour = group),size = 0.5, position = position_jitter(width = 0.33, height = 0)) +
    geom_boxplot(fill = NA,notch = FALSE, na.rm = TRUE, outlier.color = NA) +
    facet_grid(Sample.Source ~ celltype,switch = "y" ) +
    scale_colour_manual(values = c("#B61927", "#4F8FC4")) +
    guides(fill = FALSE, colour = FALSE) +
    theme(
      panel.background = element_rect(fill = "white"),
      strip.background = element_blank(),
      strip.switch.pad.grid = unit(0,units ="pt"),
      panel.spacing.x = unit(0,units = "pt"), 
      strip.placement = "outside",
      panel.border = element_blank(),
      strip.text.x = element_text(face = "bold", hjust = 0.5, vjust = 1, margin = margin(t =5, b = 5, unit = "pt")),
      strip.text.y.left = element_text(angle = 0),
      plot.title = element_text(face = "bold"),
      plot.margin = margin()
    ) +
    labs(x = "Sample Group", y = "Estimated Proportion", title = "Cell type estimations in Cortex from Cortex reference") +
    scale_y_continuous(labels = scales::percent_format(accuracy = 1), limits = c(0,1)) + 
    geom_text(nudge_x = 0.5, data = All_with_meta[All_with_meta$Sample.Source %in% c("Cortex_Frontal", "Cortex_Motor"), ], aes(x = 1, y = 0.9, label = padj_label) ) + theme_bw()
ggsave(plot = plot_all, filename =  here::here("motor_frontal_cortex.jpeg"), width = 6, height = 3, dpi = 300)
```

Information about the total number of samples here

```{r}
sample_count <- ALL %>%
  filter(!duplicated(ALL$sampleID)) %>%
  group_by(group) %>%
  summarise(count = n())
```

plot the overall box plot for each cell type here

```{r}
plot_all_cortex <- All_with_meta %>%
  ggplot(aes(x = group, proportion)) +
      geom_point(aes(colour = group),size = 0.5, position = position_jitter(width = 0.33, height = 0)) +
    geom_boxplot(fill = NA,notch = FALSE, na.rm = TRUE, outlier.color = NA) +
    facet_grid( ~ celltype,switch = "y" ) +
    scale_colour_manual(values = c("#B61927", "#4F8FC4")) +
    guides(fill = FALSE, colour = FALSE) +
    theme(
      strip.background = element_blank(),
      strip.switch.pad.grid = unit(0,units ="pt"),
      panel.spacing.x = unit(0,units = "pt"), 
      strip.placement = "outside",
      panel.border = element_blank(),
      strip.text.x = element_text(face = "bold", hjust = 0.5, vjust = 1, margin = margin(t =5, b = 5, unit = "pt")),
      strip.text.y.left = element_text(angle = 0),
      plot.margin = margin()
    ) +
    labs(x = "Sample Group", y = "Estimated Proportion", title = "Overall Cell Type Estimations in Cortex") +
    scale_y_continuous(labels = scales::percent_format(accuracy = 1), limits = c(0,1)) + 
    geom_text(nudge_x = 0.5, data = p_df, aes(x = 1, y = 0.9, label = padj) ) + theme_bw()
ggsave(plot = plot_all_cortex, filename =  here::here("ALL_cortex_all_atonce.jpeg"), width = 10, height = 3.5)
```