---
title: "Single_Cell_Cortex and Bulk Cortex Deconvolution"
author: "Nischal Bhandari"
date: "7/11/2023"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
cortex_single <- readRDS("cortex_single_seurat.rds")
```

Deconvolution Using Cortex Sample

```{r}
library(MuSiC)
library(MuSiC2)
library(SCDC)
library(SummarizedExperiment)
library(tidyr)
library(dplyr)
library(ggplot2)
```

Relabelling the cell types
```{r}
cortex_single@meta.data <- cortex_single@meta.data %>%
  mutate(broad.cell.type = case_when(
     broad.cell.type == "Oli" ~ "Oligodendrocytes",
    broad.cell.type %in% c("In", "Ex") ~ "Neurons",
    broad.cell.type == "Mic" ~ "Microglia",
    broad.cell.type == "Ast" ~ "Astrocytes",
    broad.cell.type == "End" ~ "Endothelial",
    broad.cell.type == "Per" ~ "Pericytes",
    TRUE ~ broad.cell.type 
  ))
```



```{r}
# changing the single cell SO to SCE
singlecortex.sce <- as.SingleCellExperiment(cortex_single)
```
```{r}
# exploring single cell data
p1 <- DimPlot(cortex_single, reduction = 'umap', group.by = "broad.cell.type", label = TRUE)
p1 <- p1 + theme(legend.position = "NONE") +ggtitle("Cell types in cortex single cell reference")
ggsave(p1, filename = "umap_cortex.jpeg", dpi = 300, height = 4, width = 6)
```

```{r}
count_cell_type <- cortex_single@meta.data %>%
  group_by(broad.cell.type) %>%
  summarize(count = n())
count_cell_type[order(count_cell_type$count, decreasing = TRUE), ]
```
Reading bulk data

```{r}

spinalled <- readRDS("bulk_spinal_expression_matrix.Rds")
total_spinal_samples <- 928
total_spinal_genes <- 34863


final_spinalled_ematrix <- exprs(spinalled) #exprs(final_spinalled_ematrix)
colnames(final_spinalled_ematrix) <- gsub("[.]", "-", colnames(final_spinalled_ematrix))

```




```{r}
meta_data <- read.csv("metadata.csv")

als_mnd <- meta_data[meta_data$Subject.Group == "ALS Spectrum MND", ]
als_mnd_sample_names <- als_mnd$X

control <-  meta_data[meta_data$Subject.Group == "Non-Neurological Control", ]
control_sample_names <- control$X
```


```{r}

# bulk_control <- final_spinalled_ematrix[, colnames(final_spinalled_ematrix) %in% control_sample_names]
bulk_control <- final_spinalled_ematrix[, colnames(final_spinalled_ematrix) %in% control_sample_names]
bulk_case <- final_spinalled_ematrix[, colnames(final_spinalled_ematrix) %in% als_mnd_sample_names]
```

# Estiimating with the MUSIC

```{r}
# MuSiC
bulk_control_healthy <- final_spinalled_ematrix
bulk_control_healthy <- bulk_control_healthy[, colnames(bulk_control_healthy) %in% c(control_sample_names, als_mnd_sample_names)]
```

```{r}
bulk_control_healthy <- na.omit(bulk_control_healthy)
bulk_control_healthy <- as.matrix(bulk_control_healthy)
prop_music=music_prop(bulk.mtx = bulk_control_healthy, sc.sce = singlecortex.sce,
                      clusters = 'broad.cell.type', samples = 'TAG', verbose = F)


saveRDS(prop_music, "CortexReferenceSpinalEstimation.rds")
```

```{r}
jitter.fig = Jitter_Est(list(data.matrix(prop_music$Est.prop.weighted),
                             data.matrix(prop_music$Est.prop.allgene)),
                        method.name = c('MuSiC', 'NNLS'), title = 'Jitter plot of Est Proportions in Spinal Samples from Cortex Reference')
jitter.fig


```

```{r}
music_predicted <- prop_music$Est.prop.weighted
nnls_predicted <- prop_music$Est.prop.allgene


 
prop_all_m = cbind('proportion'=c(music_predicted),'celltype'=rep(colnames(music_predicted), each=nrow(music_predicted)), 'sampleID'=rep(rownames(music_predicted),times=ncol(music_predicted)),'Method'='MuSiC')
prop_all_n = cbind('proportion'=c(nnls_predicted),'celltype'=rep(colnames(nnls_predicted), each=nrow(nnls_predicted)), 'sampleID'=rep(rownames(nnls_predicted),times=ncol(nnls_predicted)),'Method'='NNLS')

prop_all_m=as.data.frame(prop_all_m)
prop_all_n=as.data.frame(prop_all_n)
prop_all_m$proportion=as.numeric(as.character(prop_all_m$proportion))
prop_all_n$proportion=as.numeric(as.character(prop_all_n$proportion))
ALL=rbind(prop_all_m)
ALL$group = ifelse(ALL$sampleID %in% control_sample_names, 'Healthy', 'ALS')
```

```{r}
# validating the total number of samples in Healthy vs controls
sample_count <- ALL %>%
  group_by(group) %>%
  summarise(count = n())
# divide by 7. As there are 7 different cell types

```

Calculating P-Values

```{r}
# Subsetting to spinal samples alone

meta_data_spinal <- subset(meta_data, grepl("spinal", Sample.Source, ignore.case = TRUE))
spinal_samples <- meta_data_spinal$X

ALL <- ALL[ALL$sampleID %in% spinal_samples, ]
ALL$p_nom <- pairwise.wilcox.test(x = ALL$proportion, g = ALL$group)$p.value[1]
```

Calculating per cell type

```{r}
# ALL <- ALL[!ALL$celltype == 'Opc', ]
ALL <- ALL %>%
  mutate(celltype = case_when(
    celltype == "Oli" ~ "Oligodendrocytes",
    celltype %in% c("In", "Ex") ~ "Neurons",
    celltype == "Mic" ~ "Microglia",
    celltype == "Ast" ~ "Astrocytes",
    celltype == "End" ~ "Endothelial",
    celltype == "Per" ~ "Pericytes",
    TRUE ~ celltype 
  ))
```
dimensions of ALL: > dim(ALL)
[1] 5236    6

```{r}
# linear model to account fo other factors
meta_data_spinal <- meta_data_spinal[match(ALL$sampleID, meta_data_spinal$X), ]

All_with_meta <- cbind(ALL, meta_data_spinal)
All_with_meta$X <- NULL

# it might be over fitting
# mod <- lm( proportion ~ factor(Project) + factor(Site.Specimen.Collected) + factor(Family.History.of.ALS.FTD.)+ factor(Site.of.Motor.Onset) + factor(Sex) + factor(Sample.Source) + factor(Prep) + factor(Platform) + factor(Ethnicity), All_with_meta)
mod <- lm( proportion ~ factor(Site.Specimen.Collected) +  factor(Platform), All_with_meta)

All_with_meta$resid <-  residuals(mod)
```
 Calculating p - values [one normal wilcox paired-test, non-parametric, and other after regression modle--which will be shown in the graphs]
 

```{r}
# Create an empty list to store the p-values for each cell type
p_nom <- list()
p_resid <- list()
# Get unique cell types from the 'celltype' column
unique_celltypes <- unique(All_with_meta$celltype)

# Iterate over each unique cell type
for (cell in unique_celltypes) {
  # Subset the data for the current cell type
  subset_data <- subset(All_with_meta, celltype == cell)
  
  # Perform pairwise Wilcoxon test for the current cell type
  pairwise_result <- pairwise.wilcox.test(x = subset_data$proportion, g = subset_data$group)
  residual_result <- pairwise.wilcox.test(x = subset_data$resid, g = subset_data$group)
  # Extract the p-values and store them in the list
  p_nom[[cell]] <- pairwise_result$p.value[1]
  p_resid[[cell]] <- residual_result$p.value[1]
}

All_with_meta$p_nom <- NULL
All_with_meta$p_resid <- NULL

# store the p-values
p_values_df <- data.frame(celltype = names(p_nom), p_nom = unlist(p_nom), p_resid = unlist(p_resid), stringsAsFactors = FALSE)

# Merge the p_values_df with the ALL data frame
All_with_meta <- merge(All_with_meta, p_values_df, by = "celltype", all.x = TRUE)

```

plotting the box_plot and p_Values

```{r}

library(ggplot2)

# the boxplot
plot <- ggplot(All_with_meta, aes(x = celltype, y = proportion, fill = group)) +
  geom_boxplot(outlier.colour = NA) +
  labs(x = "Cell Type", y = "Estimated Proportion", fill = "Group", title = "Estimated Cell Type Proportions with Cortex Reference")

# Add p-values as text above each cell type
plot <- plot +
  geom_text(data = All_with_meta[!duplicated(ALL$celltype), ],
            aes(label = paste0("p = ", signif(p_resid, 3))),
            position = position_dodge(width = 0.35), vjust = -20) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1), limits = c(0,1))

# Adjust the plot appearance
plot <- plot +
  theme_bw() +
  theme(legend.position = "right", 
        axis.text.x = element_text(angle = 45, hjust = 1),
        plot.title = element_text(face = "bold"))

# Display the plot
plot

```
```{r}
ggsave(plot = plot,"Estimation_with_p_value_all_cell_type.jpeg", height = 8, width = 8)
write.csv(All_with_meta, "final_spinal_estimations_with_p_values.csv")
```

Plotting the graphs for cortex cell estimations with p-values

```{r}
  
p_df <- All_with_meta %>% 
    group_by(celltype, Sample.Source) %>% 
    summarise(p_nom = min(p_nom), p_resid = min(p_resid)) %>%
    mutate(padj = p.adjust(p_resid, method = "bonferroni")) %>%
    mutate(padj_label = case_when(
    padj < 0.0001 ~ "***",
    padj < 0.001 ~ "**",
    padj < 0.05 ~ "*",
    TRUE ~ "."
    ))
P_df_subset <- p_df[, c("celltype", "padj", "padj_label", "Sample.Source")]
All_with_meta <- merge(All_with_meta, P_df_subset, by = c("celltype", "Sample.Source"), all.x = TRUE)

All_with_meta <- All_with_meta[All_with_meta$Method == "MuSiC", ]
  
```

Saving all faceted image
```{r}
# calculating the median estimated proportions for each cell types

median_prop <- All_with_meta %>%
  group_by(celltype) %>%
  summarise(median_prop = median(proportion))
```
# removing opcs as they are 0

```{r}
All_with_meta <- All_with_meta[!All_with_meta$celltype == "Opc", ]
# removing spinal Unspecified as well
All_with_meta <- All_with_meta[!All_with_meta$Sample.Source == "Spinal_Cord_Unspecified ", ]
```
```{r}
plot_all <- All_with_meta %>%
  ggplot(aes(x = group, y = proportion, colour = group)) +
  geom_point(size = 0.5, position = position_jitter(width = 0.33, height = 0)) +
  geom_boxplot(notch = FALSE, na.rm = TRUE, outlier.color = NA) +
  facet_grid(Sample.Source ~ celltype, switch = "y") +
  scale_colour_manual(values = c("#B61927", "#4F8FC4")) + 
  guides(fill = FALSE, colour = guide_legend(override.aes = list(size = 2))) +
  theme(
    panel.background = element_rect(fill = "white"),
    strip.background = element_blank(),
    strip.switch.pad.grid = unit(0, units = "pt"),
    panel.spacing.x = unit(0, units = "pt"), 
    strip.placement = "outside",
    panel.border = element_blank(),
    strip.text.x = element_text(face = "bold", hjust = 0.5, vjust = 1, margin = margin(t = 5, b = 5, unit = "pt")),
    strip.text.y.left = element_text(angle = 0),
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank(),
    legend.title = element_blank(),  # Remove the legend title
    legend.position = "top",  # Position the legend at the top
    legend.direction = "horizontal",  # Display the legend horizontally
    legend.spacing = unit(0, "pt"),  # Remove spacing between legend items
    plot.margin = margin()
  ) +
  labs(x = "", y = "Estimated Proportion", title = "Cell Type Estimations in Spinal From Cortex Reference") +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1), limits = c(0, 1)) + 
  geom_text(nudge_x = 0.5, data = All_with_meta, aes(x = 1, y = 0.9, label = padj_label), color = "black")
ggsave(plot = plot_all, filename =  here::here("ALL_spinal_no_unspecified.jpeg"), width = 8, height = 9)
```

Information about the total number of samples here

```{r}
sample_count <- ALL %>%
  filter(!duplicated(ALL$sampleID)) %>%
  group_by(group) %>%
  summarise(count = n())

# after dropping Spinal Cord Unspecified

sample_count_no_unspecified <- All_with_meta %>%
  filter(!duplicated(All_with_meta$sampleID)) %>%
  group_by(group) %>%
  summarise(count = n())
```

plot the overall box plot for each cell type here

```{r}
plot_all_spinal <- All_with_meta %>%
  ggplot(aes(x = group, y = proportion, colour = group)) +
  geom_point(size = 0.5, position = position_jitter(width = 0.33, height = 0)) +
  geom_boxplot(notch = FALSE, na.rm = TRUE, outlier.color = NA, alpha = 0.0) +
  facet_grid(~ celltype, switch = "x") +
  scale_colour_manual(values = c("#B61927", "#4F8FC4")) + 
  guides(fill = FALSE, colour = guide_legend(override.aes = list(size = 2))) +
  theme(
    panel.background = element_rect(fill = "white"),
    strip.background = element_blank(),
    strip.switch.pad.grid = unit(0, units = "pt"),
    panel.spacing.x = unit(0, units = "pt"), 
    strip.placement = "outside",
    panel.border = element_blank(),
    strip.text.x = element_text(face = "bold", hjust = 0.5, vjust = 1, margin = margin(t = 5, b = 5, unit = "pt")),
    strip.text.y.left = element_text(angle = 0),
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank(),
    legend.title = element_blank(),  # Remove the legend title
    legend.position = "top",  # Position the legend at the top
    legend.direction = "horizontal",  # Display the legend horizontally
    legend.spacing = unit(0, "pt"),  # Remove spacing between legend items
    plot.margin = margin()
  ) +
  labs(x = "Cell Type", y = "Estimated Proportion", title = "Cell Type Estimations in Spinal From Cortex Reference") +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1), limits = c(0, 1)) + 
  geom_text(nudge_x = 0.5, data = All_with_meta, aes(x = 1, y = 0.9, label = signif(padj, 3)), color = "black")
plot_all_spinal <- plot_all_spinal + theme(plot.title = element_text(face = "bold"))

ggsave(plot = plot_all_spinal, filename =  here::here("ALL_spinal_all_atonce.jpeg"), width = 10, height = 3.5)
```


Combining the meta_Data to look for covariates (platform and gender)

```{r}
bulk_spinal <- read.csv("final_spinal_estimations_with_p_values.csv")
bulk_spinal$X <- NULL
bulk_spinal <- subset(bulk_spinal, sampleID %in% c(als_mnd_sample_names, control_sample_names))
bulk_spinal <- bulk_spinal[bulk_spinal$celltype != "Opc", ]
```

```{r}
# calculating p-values for platform
# Create an empty list to store the p-values for each cell type
p_nom <- list()
# Get unique cell types from the 'celltype' column
unique_celltypes <- unique(bulk_spinal$celltype)

# Iterate over each unique cell type
for (cell in unique_celltypes) {
  # Subset the data for the current cell type
  subset_data <- subset(bulk_spinal, celltype == cell)
  
  # Perform pairwise Wilcoxon test for the current cell type
  pairwise_result <- pairwise.wilcox.test(x = subset_data$proportion, g = subset_data$Platform)
  # Extract the p-values and store them in the list
  p_nom[[cell]] <- pairwise_result$p.value[1]
}


# store the p-values
p_values_df <- data.frame(celltype = names(p_nom), p_nom = unlist(p_nom), stringsAsFactors = FALSE)
```
Platform 

```{r}
plot_platform <- bulk_spinal %>%
  ggplot(aes(x = Platform, y = proportion, colour = Platform)) +
  geom_point(size = 0.5, position = position_jitter(width = 0.33, height = 0)) +
  geom_boxplot(notch = FALSE, na.rm = TRUE, outlier.color = NA, alpha = 0.3) +
  facet_grid( ~ celltype, switch = "y") +
  scale_colour_manual(values = c("#B61927", "#4F8FC4")) + 
  guides(fill = FALSE, colour = guide_legend(override.aes = list(size = 2))) + labs(x = "", y = "Estimated Proportion", title = "Estimations differences across sequencing platforms") +
  theme(plot.title = element_text(face = "bold"),
    panel.background = element_rect(fill = "white"),
    strip.background = element_blank(),
    strip.switch.pad.grid = unit(0, units = "pt"),
    panel.spacing.x = unit(0, units = "pt"), 
    strip.placement = "outside",
    panel.border = element_blank(),
    strip.text.x = element_text(face = "bold", hjust = 0.5, vjust = 1, margin = margin(t = 5, b = 5, unit = "pt")),
    strip.text.y.left = element_text(angle = 0),
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank(),
    legend.title = element_blank(),  
    legend.position = "top",
    legend.direction = "horizontal",
    legend.spacing = unit(0, "pt"),
    plot.margin = margin()
  ) + 
  scale_y_continuous(labels = scales::percent_format(accuracy = 1), limits = c(0, 1)) + 
  geom_text(nudge_x = 0.5, data = p_values_df, aes(x = 1, y = 0.9, label = signif(p_nom, 3)), color = "black")
ggsave(plot = plot_platform, "platforms_estimations.jpeg", dpi = 300, height = 3, width= 6)
```

For gender

```{r}
# calculating p-values for gender
# delete previous lists
p_nom <- list()
# Get unique cell types from the 'celltype' column
unique_celltypes <- unique(bulk_spinal$celltype)

# Iterate over each unique cell type
for (cell in unique_celltypes) {
  # Subset the data for the current cell type
  subset_data <- subset(bulk_spinal, celltype == cell)
  
  # Perform pairwise Wilcoxon test for the current cell type
  pairwise_result <- pairwise.wilcox.test(x = subset_data$proportion, g = subset_data$Sex)
  # Extract the p-values and store them in the list
  p_nom[[cell]] <- pairwise_result$p.value[1]
}


# store the p-values
p_values_df <- data.frame(celltype = names(p_nom), p_nom = unlist(p_nom), stringsAsFactors = FALSE)
```

```{r}
plot2 <- bulk_spinal %>%
  ggplot(aes(x = Sex, y = proportion, colour = Sex)) +
  geom_point(size = 0.5, position = position_jitter(width = 0.33, height = 0)) +
  geom_boxplot(notch = FALSE, na.rm = TRUE, outlier.color = NA, alpha = 0.3) +
  facet_grid( ~ celltype, switch = "y") +
  scale_colour_manual(values = c("#B61927", "#4F8FC4")) + 
  guides(fill = FALSE, colour = guide_legend(override.aes = list(size = 2))) + labs(x = "", y = "Estimated Proportion", title = "Estimations differences across gender") +
  theme(plot.title = element_text(face = "bold"),
    panel.background = element_rect(fill = "white"),
    strip.background = element_blank(),
    strip.switch.pad.grid = unit(0, units = "pt"),
    panel.spacing.x = unit(0, units = "pt"), 
    strip.placement = "outside",
    panel.border = element_blank(),
    strip.text.x = element_text(face = "bold", hjust = 0.5, vjust = 1, margin = margin(t = 5, b = 5, unit = "pt")),
    strip.text.y.left = element_text(angle = 0),
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank(),
    legend.title = element_blank(),  
    legend.position = "top",
    legend.direction = "horizontal",
    legend.spacing = unit(0, "pt"),
    plot.margin = margin()
  ) + 
  scale_y_continuous(labels = scales::percent_format(accuracy = 1), limits = c(0, 1)) + 
  geom_text(nudge_x = 0.5, data = p_values_df, aes(x = 1, y = 0.9, label = signif(p_nom, 3)), color = "black")
ggsave(plot = plot2, "Cortex_Spinal/gender_estimations.jpeg", dpi = 300, height = 3, width= 6)
```

Age at symptom onset

```{r}
correlation_coefficient <- cor(bulk_spinal$Age.at.Death, bulk_spinal$Age.at.Symptom.Onset, use = "complete.obs")
p_value <- wilcox.test(bulk_spinal$Age.at.Death, bulk_spinal$Age.at.Symptom.Onset, paired = TRUE, alternative = "two.sided")
p_value <- 2.2e-16

plot3 <- ggplot(bulk_spinal, aes(Age.at.Symptom.Onset, Age.at.Death)) +
  geom_point(color = "red", size = 1) + geom_smooth(method = "lm") + theme_classic() + labs(x = "Age at symptom onset", y = "Age at death", title = "Age at death vs symptom onset") + theme(plot.title = element_text(face = "bold"))

```

```{r}

plot3 <- plot3 +
  annotate(
    "text", x = 30, y = 75,
    label = paste("Pearson's r =", round(correlation_coefficient, 3)),
    size = 4, hjust = 0, vjust = 0
  ) +  annotate(
    "text", x = 30, y = 70,
    label = paste("p-value =", p_value),
    size = 4, hjust = 0, vjust = 0)
ggsave(plot = plot3, "age_corr.jpeg", dpi = 300, height = 3, width= 4)

```

```{r}
correlation_coefficient <- cor(bulk_spinal$Age.at.Death, bulk_spinal$Disease.Duration.in.Months, use = "complete.obs")
p_value <- wilcox.test(bulk_spinal$Age.at.Death, bulk_spinal$Disease.Duration.in.Months, paired = TRUE, alternative = "two.sided")
p_value <- p_value$p.value

plot4 <- ggplot(bulk_spinal, aes(Age.at.Symptom.Onset, Disease.Duration.in.Months)) +
  geom_point(color = "red", size = 1) + geom_smooth(method = "lm") + theme_classic() + labs(x = "Age at symptom onset", y = "Disease duration (in months)", title = " Disease duration vs age at symptom onset") + theme(plot.title = element_text(face = "bold"))


plot4 <- plot4 +
  annotate(
    "text", x = 30, y = 214,
    label = paste("Pearson's r =", round(correlation_coefficient, 3)),
    size = 4, hjust = 0, vjust = 0
  ) +  annotate(
    "text", x = 30, y = 200,
    label = paste("p-value =", signif(p_value, 3)),
    size = 4, hjust = 0, vjust = 0)


ggsave(plot = plot4, "age_dur.jpeg", dpi = 300, height = 4, width= 5)
```